---
title: "Figure R0 for SIR"
author: "Subhash Lele"
date: '2022-12-16'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This constains program for constructing figure for the profile likelihood of R0 under the SIR model for the influenza cases data. 


```{r, message=FALSE}
 
library(here)
source(here("Source functions and libraries/Required libraries.R"))
source(here("Source functions and libraries/Source function for DC Profile likelihood_2.R"))
source(here("SIR model/SIR simulator.R"))
source(here("SIR model/Source functions for ML fitting of SIR.R"))
rstan_options (auto_write = TRUE)
options (mc.cores = parallel::detectCores ())

```
Data set
```{r}
cases <- c(1,6,26,73,222,293,258,236,191,124,69,26,11,4)
cases = cases[-1]   # Initiating at 1 in the data
cases1 <- array(cases,c(length(cases),1))
```

Step 2: Create the appropriate  R objects from the data set.
```{r}
N = 763
n_days <- length(cases)
t <- seq(0, n_days, by = 1)
t0 = 0
t <- t[-1]
#initial conditions
i0 <- 1
s0 <- N - i0
r0 <- 0
y0 = c(S = s0, I = i0, R = r0)
```
Step 3: Compute the MLE and its variance using data cloning
For this particular example, we do not present the full MLE calculation using data cloning. We simply use 10 clones instead of iterating through increasing number of clones and checking on the convergence of posterior variance to 0. We did that separately.

```{r}
nclones <- 10
cases1 <- dclone(dcdim(cases1),nclones)

# Fitting Poisson errors model 

MuP = rep(0,2)
SigmaP = diag(10,2,2)
niter = 2000
data_sir_MN <- list(n_days = n_days, y0 = y0, t0 = t0, ts = t, N = N, nclones=nclones,cases = cases1,Mu=MuP,Sigma=SigmaP)
unchanged=c("Mu","Sigma","n_days","y0","t0","ts","N")

# Compile and fit the model
model <- stan_model(here("SIR model/sir_poisson_MN.stan"))
fit_sir_poisson_MN <- sampling(model,
data = data_sir_MN,
iter = niter,
chains = 1)
SIR.out = extract(fit_sir_poisson_MN)  

# The MLE of the parameters
DC.MLE = apply(SIR.out$parms,2,mean)
DC.Var = nclones * var(SIR.out$parms)
```
Algorithm 2: We compute the posterior distribution with the ML asymptotic distribution as the prior distribution

```{r}
# Replace the MuP and SigmaP in the data object and rerun the program with nclones=1 and nclones=2. This will be the output needed for the profile likelihood computations. We can then compute the profile likelihood for R0 easily. 
cases1 = array(cases1,c(n_days,1))
cases <- dclone(dcdim(cases1),1)
data_sir_MN_1 <- list(n_days = n_days, y0 = y0, t0 = t0, ts = t, N = N, nclones=1,cases = cases,Mu=as.vector(DC.MLE),Sigma=as.matrix(DC.Var))

fit_sir_Poisson_MN_1 <- sampling(model,
data = data_sir_MN_1,iter = niter,chains = 1)

# Two clones
cases <- dclone(dcdim(cases1),2)
data_sir_MN_2 <- list(n_days = n_days, y0 = y0, t0 = t0, ts = t, N = N, nclones=2,cases = cases,Mu=DC.MLE,Sigma=DC.Var)

fit_sir_Poisson_MN_2 <- sampling(model,
data = data_sir_MN_2,
iter = niter,
chains = 1)

# MCMC output for computing profile likelihood.

M1 = extract(fit_sir_Poisson_MN_1)$parms
M2 = extract(fit_sir_Poisson_MN_2)$parms
grid.length=100

# Parameter of interest calculation function
psi.fn_R0 = function(M){
  psi = exp(M[,1])/exp(M[,2])
  return(psi)
}
```
Compute the profile likelihood for R0 using Algorithm 2
```{r}
DD.profile.q = profile_fn(psi.fn=psi.fn_R0,M1,M2,method="c")
psi.grid = DD.profile.q$Profile.out$psi.grid
```
Now we compute the profile likelihood using the data doubling but with flat priors. Notice that the model can be rewritten so that R0 is a marginal parameter. 
```{r}

# One clone
nclones <- 1
cases <- dclone(dcdim(cases1),nclones)
data_sir_Bayes_P <- list(n_days = n_days, y0 = y0, t0 = t0, ts = t, N = N, nclones=nclones,cases = cases)

# Compile the model with flat priors 
model <- stan_model(here("SIR model/sir_poisson_Bayes.stan"))
fit_sir_poisson_Bayes <- sampling(model,
data = data_sir_Bayes_P,iter = niter,chains = 1)
SIR.out1 = extract(fit_sir_poisson_Bayes) 


nclones <- 2
cases <- dclone(dcdim(cases1),nclones)
data_sir_Bayes_P <- list(n_days = n_days, y0 = y0, t0 = t0, ts = t, N = N, nclones=nclones,cases = cases)

fit_sir_poisson_Bayes <- sampling(model,
data = data_sir_Bayes_P,iter = niter,chains = 1)
SIR.out2 = extract(fit_sir_poisson_Bayes)

psi.R0 = function(M){
  return(M)
}

DD.Bayes.R0 = profile_fn(psi.fn=psi.R0,M1=SIR.out1$R0,M2=SIR.out2$R0,psi.grid=psi.grid,method="c")

```

Now compute the analytical profile likelihood

```{r}
# Program for analytical MLE and profile MLE for SIR model

sir_1 <- function(beta, gamma, S0, I0, R0, times) {
  require(deSolve) # for the "ode" function
  
  # the differential equations:
  sir_equations <- function(time, variables, parameters) {
    with(as.list(c(variables, parameters)), {
      dS <- -(beta) * I * S
      dI <-  (beta) * I * S - gamma * I
      dR <-  gamma * I
      return(list(c(dS, dI, dR)))
    })
  }
  # the parameters values:
  parameters_values <- c(beta  = beta, gamma = gamma)
  
  # the initial values of variables:
  initial_values <- c(S = S0, I = I0, R = R0)
  
  # solving
  out <- ode(initial_values, times, sir_equations, parameters_values)
  
  # returning the output:
  as.data.frame(out)
}

mLL_pois <- function(beta1, gamma1, day, cases) {
  beta <- exp(beta1) # to make sure that the parameters are positive
  gamma <- exp(gamma1)
  #  sigma <- exp(sigma)
  I0 <- cases[1] # initial number of infectious
  observations <- cases[-1] # the fit is done on the other data points
  predictions <- sir_1(beta = beta, gamma = gamma,
                       S0 = N - I0, I0 = I0, R0 = 0, times = day)
  predictions <- predictions$I[-1] # removing the first point too
  if (any(predictions < 0)) return(NA) # safety
  # returning minus log-likelihood:
  #  -sum(dnorm(x = observations, mean = predictions, sd = sigma, log = TRUE))
  return(-sum(dpois(x = observations, lambda = predictions, log = TRUE)))
}

flu <- read.table("https://bit.ly/2vDqAYN", header = TRUE)

starting_param_val <- list(beta1 = 0.004, gamma1 = 0.5)
estimates_pois <- bbmle::mle2(minuslogl = mLL_pois,
                     start = lapply(starting_param_val,log),
                     data = c(flu, N = 763))

beta.hat = estimates_pois@coef[1]
gamma.hat = estimates_pois@coef[2]
R0 =(exp(beta.hat)*763)/exp(gamma.hat)

```

We rewrite the function so that R0 is a marginal 
parameter to get the analytical profile likelihood.

```{r}
mLL_pois_R0 <- function(BR01, gamma1, day, cases) {
  beta1 <- log(exp(BR01)*exp(gamma1)/N)
  beta <- exp(beta1) # to make sure that the parameters are positive
  gamma <- exp(gamma1)
  I0 <- cases[1] # initial number of infectious
  observations <- cases[-1] # the fit is done on the other data points
  predictions <- sir_1(beta = beta, gamma = gamma,
                       S0 = N - I0, I0 = I0, R0 = 0, times = day)
  predictions <- predictions$I[-1] # removing the first point too
  if (any(predictions < 0)) return(NA) # safety
  # returning minus log-likelihood:
  #  -sum(dnorm(x = observations, mean = predictions, sd = sigma, log = TRUE))
  return(-sum(dpois(x = observations, lambda = predictions, log = TRUE)))
}


starting_param_val <- list(BR01 = 4, gamma1 = 0.5)
estimates_pois <- bbmle::mle2(minuslogl = mLL_pois_R0,
                     start = lapply(starting_param_val,log),
                     data = c(flu, N = 763))


R0.hat = exp(estimates_pois@coef[1])
# This should be nearly the same as the one obtained before.
c(R0,R0.hat)
```

```{r}
R0.grid = log(psi.grid)
N.grid = length(R0.grid)
out = matrix(0,N.grid,3)


for (i in 1:N.grid){
  tryCatch({
  estimates_pois.fix <- bbmle::mle2(minuslogl = mLL_pois_R0,
                                    start = lapply(starting_param_val,log),fixed=list(BR01=R0.grid[i]),
                                    data = c(flu, N = 763))},error = function(e) e)
 
  out[i,] = c(R0.grid[i],estimates_pois.fix@coef, estimates_pois.fix@min)
}

Profile.ana = cbind(exp(out[,1]),-out[,3])

# Need to standardize this.
Profile.ana[,2] = Profile.ana[,2] - max(Profile.ana[,2])

```

Get the data for the profile likelihood.

```{r}

loglike.data = data.frame(psi.grid=psi.grid,
DD.PL.q = DD.profile.q$Profile.out[,2],
DD.PL.B = DD.Bayes.R0$Profile.out[,2],
DD.PL.ana = Profile.ana[,2])

write.table(loglike.data,here("SIR model/Figure2 data.csv"))

```
Plot the profile likelihood estimates (Algorithms 1 and 2 and analytical) 

```{r}
loglike.data = read.table(here("SIR model/Figure2 data.csv"))
fig2 = ggplot() +
  geom_line(data=loglike.data,aes(x=psi.grid,y=DD.PL.q),col="black",linetype="dashed") +
  geom_line(data=loglike.data,aes(x=psi.grid,y=DD.PL.B),col="black",linetype="twodash") +
  geom_line(data=loglike.data,aes(x=psi.grid,y=DD.PL.ana),col="black") +
labs(x="Basic Reproduction Rate", y="Profile LL") +
  ylim(-3,1) + 
  theme(text = element_text(size = 6))+
  theme(plot.title = element_text(hjust = 0.5))
ggsave(here("SIR model/Figure2.pdf"),fig2,device="pdf")
```
