---
title: "Profile likelihood examples"
subtitle: "Beta distribution: Mean, Variance and Skewness"
author: "Subhash Lele"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F)
```
# Introduction

We will compute the profile likelihood for various parameters of interest for the univariate Beta distribution. 

## Global R functions and libraries:

We will source the appropriate general set of functions and libraries in the following. These are common to all models. 

```{r, message=FALSE}
library(here)
source(here("Source functions and libraries/Required libraries.R"))
source(here("Source functions and libraries/Source function for DC Profile likelihood_2.R"))
source(here("Source functions and libraries/Source function for analytical profile likelihood.R"))
```

 We will then source the model specific R functions to compute the MLE of the parameters and generate the MCMC samples based on the original and doubled data.
```{r}
source(here("Beta distribution/Functions for MLE and Profile Sampling.R"))
source(here("Beta distribution/Functions for Analytical MLE and profile likelihood.R"))
```

Now we write functions that compute the parameter of interest that is simply a function of the canonical parameters. 

```{r}

Beta.mean = function(M){
  a = exp(M[,1])
  b = exp(M[,2])
  psi = a/(a+b)
  return(psi)
}

Beta.var = function(M){
  a = exp(M[,1])
  b = exp(M[,2])
  num = a*b
  den1 = (a+b)^2
  den2 = (a+b+1)
  psi = num/(den1*den2)
  return(psi)
}

Beta.skew = function(M){
  a = exp(M[,1])
  b = exp(M[,2])
  num = 2*(b-a)*sqrt(a+b+1)
  den1 = (a+b+2)
  den2 = sqrt(a*b)
  psi = num/(den1*den2)
  return(psi)
}
```

Following are functions specify the constraint based on the parameter of interest for computing analytical profile likelihood. 

```{r}
Ana.Beta.mean = function(parms,Y,psi.fix){
  a = exp(parms[1])
  b = exp(parms[2])
  psi = a/(a+b)-psi.fix
  return(psi)
}

Ana.Beta.var = function(parms,Y,psi.fix){
  a = exp(parms[1])
  b = exp(parms[2])
  num = a*b
  den1 = (a+b)^2
  den2 = (a+b+1)
  psi = num/(den1*den2)-psi.fix
  return(psi)
}

Ana.Beta.skew = function(parms,Y,psi.fix){
  a = exp(parms[1])
  b = exp(parms[2])
  num = 2*(b-a)*sqrt(a+b+1)
  den1 = (a+b+2)
  den2 = sqrt(a*b)
  psi = num/(den1*den2)-psi.fix
  return(psi)
}

```
Let us now apply this on some generated data. We will start with sample size 10. 

```{r}
shape1 = 3
shape2 = 5
n = 10
Y = rbeta(n,shape1,shape2)
```
Data cloning for MLE and data doubling samples. 

```{r}
# Compute the MLE first
Y = dcdim(data.matrix(Y))
dat = list(Y=Y,n=n,ncl=1,MuP=rep(0,2),PrecP=diag(0.01,2,2))

n.clones = c(1,5,10)
params = c("parms")

DC.MLE = dc.fit(dat,params,model=DC.MLE.fn,n.clones=n.clones,multiply="ncl",unchanged=c("n","MuP","PrecP"),n.chains=5,n.update=1000,n.iter=5000,n.adapt=2000)

# Check the convergence and MLE estimates etc.
# summary(DC.MLE)
# dctable(DC.MLE)
# dcdiag(DC.MLE)

# Step 2: Generate samples using MLE distribution as the prior

dat.p = list(Y=Y,n=n,ncl=1,MLE=coef(DC.MLE),FI=solve(vcov(DC.MLE)))
unchanged=c("n","MLE","FI")
params = c("parms")

ProfileSample_A2 = DDSample_fn(DC.profile.fn,dat.p,params,unchanged,n.chains=50,n.update=1000,n.iter=1000,n.adapt=5000)

```
Now estimate the profile likelihood function, Profile MLE and associated confidence interval. We also get the grid points for applying Algorithm 1.
```{r}
DD.profile.q = profile_fn(psi.fn=Beta.mean,ProfileSample_A2$M1,ProfileSample_A2$M2,method="c")
psi.grid = DD.profile.q$Profile.out$psi.grid
```
Algorithm 1: Parameters of interest is a marginal parameters. We use data doubling with a flat prior.

```{r}
Y = dcdim(data.matrix(Y))
dat = list(Y=Y,n=n,ncl=1)
params = c("mu1")
unchanged=c("n")

ProfileSample_A1 = DDSample_fn(Bayes.fn,dat,params,unchanged,n.chains=5,n.update=1000,n.iter=1000,n.adapt=5000)

Beta.psi = function(M){
  out = M  # We only monitor the parameter of interest
  return(out)
}

DD.profile.psi = profile_fn(psi.fn=Beta.psi,ProfileSample_A1$M1,ProfileSample_A1$M2, method="c",psi.grid=psi.grid)
```

We can also compute analytical profile likelihood in this case.

```{r}
parms.start = c(log(shape1),log(shape2))
Ana.profile.mean = profile.analytical(parms.start,floglike.Beta,fpsi=Ana.Beta.mean,Y,psi.grid)
# Need to standardize this function
Ana.profile.mean$profile.like = -Ana.profile.mean$profile.like + as.numeric(min(Ana.profile.mean$profile.like))

```
Get the data for the profile likelihood

```{r}
loglike.data = data.frame(psi.grid=psi.grid,
DD.PL.q = DD.profile.q$Profile.out[,2],
DD.PL.psi = DD.profile.psi$Profile.out[,2],
Analytical.PL = Ana.profile.mean[,2])
write.table(loglike.data,here(here("Beta distribution/Figure 1a data.csv")))
```

Plot the profile likelihood
```{r}
loglike.data = read.table(here("Beta distribution/Figure 1a data.csv"))
fig1a = ggplot() +
  geom_line(data=loglike.data,aes(x=psi.grid,y=DD.PL.q),col="black",linetype="dashed") +
  geom_line(data=loglike.data,aes(x=psi.grid,y=DD.PL.psi),col="black",linetype="twodash") +
  geom_line(data=loglike.data,aes(x=psi.grid,y=Analytical.PL),col="black",linetype="solid") +
labs(x="Mean", y="Profile LL") +
  ylim(-3,1) + 
ggtitle("N=10")+
  theme(text = element_text(size = 6))+
  theme(plot.title = element_text(hjust = 0.5))

```

We will now use Algorithm 2 and analytical profile likelihood for variance and skewness parameters. Notice that Algorithm 1 is not applicable (at least not easily) in these cases. 


```{r}

#Profile likelihood for variance

DD.profile.q = profile_fn(psi.fn=Beta.var,ProfileSample_A2$M1,ProfileSample_A2$M2,method="c")
psi.grid = DD.profile.q$Profile.out$psi.grid

# Analytical profile likelihood for variance

parms.start = c(log(shape1),log(shape2))
Ana.profile.var = profile.analytical(parms.start,floglike.Beta,fpsi=Ana.Beta.var,Y,psi.grid)
# Need to standardize this function
Ana.profile.var$profile.like = -Ana.profile.var$profile.like + as.numeric(min(Ana.profile.var$profile.like))
```
Get the data for the profile likelihood 

```{r}
loglike.data = data.frame(psi.grid=psi.grid,
DD.PL.q = DD.profile.q$Profile.out[,2],
Analytical.PL = Ana.profile.var[,2])
write.table(loglike.data,here(here("Beta distribution/Figure 1c data.csv")))
```
Plot the profile likelihood
```{r}
loglike.data = read.table(here("Beta distribution/Figure 1c data.csv"))
fig1c = ggplot() +
  geom_line(data=loglike.data,aes(x=psi.grid,y=DD.PL.q),col="black",linetype="dashed") +
  geom_line(data=loglike.data,aes(x=psi.grid,y=Analytical.PL),col="black",linetype="solid") +
labs(x="Variance", y="Profile LL") +
  ylim(-3,1) + 
ggtitle("N=10")+
  theme(text = element_text(size = 6))+
  theme(plot.title = element_text(hjust = 0.5))
```

Profile likelihood for skewness

```{r}
DD.profile.q = profile_fn(psi.fn=Beta.skew,ProfileSample_A2$M1,ProfileSample_A2$M2,method="c")
psi.grid = DD.profile.q$Profile.out$psi.grid

# Analytial profile likelihood
parms.start = c(log(shape1),log(shape2))
Ana.profile.skew = profile.analytical(parms.start,floglike.Beta,fpsi=Ana.Beta.skew,Y,psi.grid)
# Need to standardize this function
Ana.profile.skew$profile.like = -Ana.profile.skew$profile.like + as.numeric(min(Ana.profile.skew$profile.like))
```
Get the data for the profile likelihood 

```{r}
loglike.data = data.frame(psi.grid=psi.grid,
DD.PL.q = DD.profile.q$Profile.out[,2],
Analytical.PL = Ana.profile.skew[,2])
write.table(loglike.data,here(here("Beta distribution/Figure 1e data.csv")))
```
Plot the profile likelihood
```{r}
loglike.data = read.table(here("Beta distribution/Figure 1e data.csv"))
fig1e = ggplot() +
  geom_line(data=loglike.data,aes(x=psi.grid,y=DD.PL.q),col="black",linetype="dashed") +
  geom_line(data=loglike.data,aes(x=psi.grid,y=Analytical.PL),col="black",linetype="solid") +
labs(x="Skewness", y="Profile LL") +
  ylim(-3,1) + 
ggtitle("N=10")+
  theme(text = element_text(size = 6))+
  theme(plot.title = element_text(hjust = 0.5))


```

We will now plot the corresponding figures for N=50 

```{r}
shape1 = 3
shape2 = 5
n = 50
Y = rbeta(n,shape1,shape2)
```
Data cloning for MLE and data doubling samples. 

```{r}
# Compute the MLE first
Y = dcdim(data.matrix(Y))
dat = list(Y=Y,n=n,ncl=1,MuP=rep(0,2),PrecP=diag(0.01,2,2))

n.clones = c(1,5,10)
params = c("parms")

DC.MLE = dc.fit(dat,params,model=DC.MLE.fn,n.clones=n.clones,multiply="ncl",unchanged=c("n","MuP","PrecP"),n.chains=5,n.update=1000,n.iter=5000,n.adapt=2000)

# Check the convergence and MLE estimates etc.
# summary(DC.MLE)
# dctable(DC.MLE)
# dcdiag(DC.MLE)

# Step 2: Generate samples using MLE distribution as the prior

dat.p = list(Y=Y,n=n,ncl=1,MLE=coef(DC.MLE),FI=solve(vcov(DC.MLE)))
unchanged=c("n","MLE","FI")
params = c("parms")

ProfileSample_A2 = DDSample_fn(DC.profile.fn,dat.p,params,unchanged,n.chains=50,n.update=1000,n.iter=1000,n.adapt=5000)

```
Now estimate the profile likelihood function, Profile MLE and associated confidence interval. We also get the grid points for applying Algorithm 1.
```{r}
DD.profile.q = profile_fn(psi.fn=Beta.mean,ProfileSample_A2$M1,ProfileSample_A2$M2,method="c")
psi.grid = DD.profile.q$Profile.out$psi.grid
```
Algorithm 1: Parameters of interest is a marginal parameters. We use data doubling with a flat prior.

```{r}
Y = dcdim(data.matrix(Y))
dat = list(Y=Y,n=n,ncl=1)
params = c("mu1")
unchanged=c("n")

ProfileSample_A1 = DDSample_fn(Bayes.fn,dat,params,unchanged,n.chains=5,n.update=1000,n.iter=1000,n.adapt=5000)

Beta.psi = function(M){
  out = M  # We only monitor the parameter of interest
  return(out)
}

DD.profile.psi = profile_fn(psi.fn=Beta.psi,ProfileSample_A1$M1,ProfileSample_A1$M2, method="c",psi.grid=psi.grid)
```

We can also compute analytical profile likelihood in this case.

```{r}
parms.start = c(log(shape1),log(shape2))
Ana.profile.mean = profile.analytical(parms.start,floglike.Beta,fpsi=Ana.Beta.mean,Y,psi.grid)
# Need to standardize this function
Ana.profile.mean$profile.like = -Ana.profile.mean$profile.like + as.numeric(min(Ana.profile.mean$profile.like))

```
Get the data for the profile likelihood

```{r}
loglike.data = data.frame(psi.grid=psi.grid,
DD.PL.q = DD.profile.q$Profile.out[,2],
DD.PL.psi = DD.profile.psi$Profile.out[,2],
Analytical.PL = Ana.profile.mean[,2])
write.table(loglike.data,here(here("Beta distribution/Figure 1b data.csv")))
```

Plot the profile likelihood
```{r}
loglike.data = read.table(here("Beta distribution/Figure 1b data.csv"))
fig1b = ggplot() +
  geom_line(data=loglike.data,aes(x=psi.grid,y=DD.PL.q),col="black",linetype="dashed") +
  geom_line(data=loglike.data,aes(x=psi.grid,y=DD.PL.psi),col="black",linetype="twodash") +
  geom_line(data=loglike.data,aes(x=psi.grid,y=Analytical.PL),col="black",linetype="solid") +
labs(x="Mean", y="Profile LL") +
  ylim(-3,1) + 
ggtitle("N=50")+
  theme(text = element_text(size = 6))+
  theme(plot.title = element_text(hjust = 0.5))

```

We will now use Algorithm 2 and analytical profile likelihood for variance and skewness parameters. Notice that Algorithm 1 is not applicable (at least not easily) in these cases. 


```{r}

#Profile likelihood for variance

DD.profile.q = profile_fn(psi.fn=Beta.var,ProfileSample_A2$M1,ProfileSample_A2$M2,method="c")
psi.grid = DD.profile.q$Profile.out$psi.grid

# Analytical profile likelihood for variance

parms.start = c(log(shape1),log(shape2))
Ana.profile.var = profile.analytical(parms.start,floglike.Beta,fpsi=Ana.Beta.var,Y,psi.grid)
# Need to standardize this function
Ana.profile.var$profile.like = -Ana.profile.var$profile.like + as.numeric(min(Ana.profile.var$profile.like))
```
Get the data for the profile likelihood 

```{r}
loglike.data = data.frame(psi.grid=psi.grid,
DD.PL.q = DD.profile.q$Profile.out[,2],
Analytical.PL = Ana.profile.var[,2])
write.table(loglike.data,here(here("Beta distribution/Figure 1d data.csv")))
```
Plot the profile likelihood
```{r}
loglike.data = read.table(here("Beta distribution/Figure 1d data.csv"))
fig1d = ggplot() +
  geom_line(data=loglike.data,aes(x=psi.grid,y=DD.PL.q),col="black",linetype="dashed") +
  geom_line(data=loglike.data,aes(x=psi.grid,y=Analytical.PL),col="black",linetype="solid") +
labs(x="Variance", y="Profile LL") +
  ylim(-3,1) + 
ggtitle("N=50")+
  theme(text = element_text(size = 6))+
  theme(plot.title = element_text(hjust = 0.5))

```

Profile likelihood for skewness

```{r}
DD.profile.q = profile_fn(psi.fn=Beta.skew,ProfileSample_A2$M1,ProfileSample_A2$M2,method="c")
psi.grid = DD.profile.q$Profile.out$psi.grid

# Analytial profile likelihood
parms.start = c(log(shape1),log(shape2))
Ana.profile.skew = profile.analytical(parms.start,floglike.Beta,fpsi=Ana.Beta.skew,Y,psi.grid)
# Need to standardize this function
Ana.profile.skew$profile.like = -Ana.profile.skew$profile.like + as.numeric(min(Ana.profile.skew$profile.like))
```
Get the data for the profile likelihood 

```{r}
loglike.data = data.frame(psi.grid=psi.grid,
DD.PL.q = DD.profile.q$Profile.out[,2],
Analytical.PL = Ana.profile.skew[,2])
write.table(loglike.data,here(here("Beta distribution/Figure 1f data.csv")))
```
Plot the profile likelihood
```{r}
loglike.data = read.table(here("Beta distribution/Figure 1f data.csv"))
fig1f = ggplot() +
  geom_line(data=loglike.data,aes(x=psi.grid,y=DD.PL.q),col="black",linetype="dashed") +
  geom_line(data=loglike.data,aes(x=psi.grid,y=Analytical.PL),col="black",linetype="solid") +
labs(x="Skewness", y="Profile LL") +
  ylim(-3,1) + 
ggtitle("N=50")+
  theme(text = element_text(size = 6))+
  theme(plot.title = element_text(hjust = 0.5))

```

Plot these in one figure

```{r}
library(cowplot)
Figure1 = plot_grid(fig1a,fig1b,fig1c,fig1d,fig1e,fig1f,ncol=2)
ggsave2(here("Beta distribution/Figure1.pdf"),Figure1,device="pdf")
```
